The Ranked Nearest Neighbour Interchange space

### 2019-06-25 paper.tex
number of ctrees with max dist from given ctree
lots of definitions

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}

@@

%Do we restrict everything to RNNI? Make sure that we correctly distinguish between RNNI and RNNIu (for instance in the algorithm section)

%ranks of internal node shall be from 1 to n, leaves have rank 0

%MRCA

%cluster

%leaf or taxon?

%caterpillar path, shortest caterpillar path, caterpillar distance (def currently in proof of lemma:caterpillar_dist=diameter)

%convexity (there is a shortest path within that set, not: all shortest path are within this set)

\begin{definition}
	Let $A$ be a pending subtree of the tree $T$.
	We call the set $c(A)$ of taxa of $A$ the \emph{cluster} induced by $A$.
\end{definition}

@@

	\textbf{Rank swaps:}
	Within rank swaps the labels of internal nodes do not change.
	For an example consider Figure~\ref{fig:rank_swap_internal_label}.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank swap.
		}
		\label{fig:rank_swap_internal_label}
	\end{figure}

@@

	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
	Therefore, we use the following observation:
	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
	This is true for both rank swaps and $\nni$ moves, according to our definition of the labelling:
	Within a rank swap two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.

@@

%This part is mainly for myself, not sure whether it is important to have it in the paper
The counterexample presented in~\cite{Li1996} works in $\nni$ is based on following observation:
Let $T_1$ and $T_2$ be two caterpillar trees.
If each taxon $a$ is replaced by a cherry $(a_1,a_2)$ in both trees $T_1$ and $T_2$, the resulting trees $T_1'$ and $T_2'$ have distance $d_{\nni}(T_1',T_2') = d_{\nni}(T_1,T_2)$ in the $\nni$ graph.
When proving~\ref{thm:split_nni},~\cite{Li1996} give an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
The split given by the edges incident to the root is $1, \ldots, n | n+1, \ldots, 2n$ in both trees $T$ and $R$.
For the given example there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $T'$ is tranformed to a tree $R'$ that contains the same cherries as $T'$.
Finally, all cherries of $R'$ are resolved in order to end up in tree $R$.
For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(T',R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ is being used.
However, this does not work in $\rnni$.
Introducing new cherries means introducing new internal nodes with new ranks.
Since there are some rank swaps needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the split $1, \ldots, n | n+1, \ldots, 2n$ is no shortest path.
% How do we assign the ranks to the new cherry nodes?
% We could either create a path where $T'$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank swaps before and after that part OR
% Build $T'$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $T'$ and $R'$.
% Also anything in between these two options is possible. Are we able to disprove every option??

@@

% idea for description of max distance caterpillar trees:

@@

According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$ and the ranks of their internal nodes.
For caterpillar trees it is $(n-1)!$.

@@

>>>>

\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{float}
\usepackage{graphicx}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{todonotes}
\usepackage{enumitem}

@@

\todo{citation of definitions??}

A \amph{rooted binary phylogenetic tree} a set $X$ is a pair $(t, \Phi)$ where $t$ is a rooted binary tree with leaf set $\mathcal{L}(t)$ and $\Phi:X \mapsto \mathcal{L}(T)$ a bijective mapping. \todo{do we need to define leaves and internal nodes, child, parent?}
A \emph{ranked tree} is given by a pair of such a rooted binary phylogenetic tree $(t, \Phi)$ on $X = \{1, \ldots, n\}$ for $n \in \mathbb N$, and a rank function that maps the vertex set of $t$ onto the set $\{0,\ldots,n-1\}$ such that all leaves are assigned rank $0$ and each internal node has rank greater than its children.
The number $n$ of taxa will remain fixed throughout the whole paper.
For simplicity of notation we use the term \emph{tree} instead instead of ranked tree.
Furthermore, we will identify leaves by their labels and might call them \emph{taxa} (singular: \emph{taxon}).
For a tree $T$ a subtree consisting of an internal node with both children being leaves is called \emph{cherry}.
An edge $e$ that has no leaf as child is an \emph{internal edge}.
The set of taxa that are descending from an internal edge $e$ is called \emph{cluster}. \todo{OK?}
In the literature, clusters often refer to sets of taxa descending from internal nodes.
The set of clusters as defined above would then be recalled as non-trivial clusters.
Given a subset $M$ of the set of taxa $X$, the \emph{most recent common ancestor} $mrca(M)$ of $M$ is the internal node with lowest rank that is ancestor of all taxa in $M$. \todo{OK?}
For a single taxon $x$, $mrca(\{x\})$ is the parent of $x$ and will be denoted by $parent(x)$.
Two trees are called \emph{isomorphic} if the underlying rooted phylogenetic trees are isomorphic and the isomorphism preserves ranks of the nodes.

The set of ranked trees builds the vertex set of the \emph{$\rnni$-graph}.
We introduce two types of operations (\emph{$\rnni$ moves}) connecting trees $T$ and $R$ in that graph by an edge:
The swap of the ranks of two internal nodes that have rank difference one is called \emph{rank move}.
We say that there is an $\nni$ move between trees $T$ and $R$ if there are edges $e$ in $T$ and $f$ in $R$ that are bounded by internal nodes with rank difference one such that the graphs obtained by shrinking $e$ and $f$ to internal nodes are isomorphic.
Trees that are connected by either a rank swap or an $\nni$ move are called \emph{$\rnni$ neighbours} or briefly neighbours.
Note that the previously defined $\rnni$ graph is a metric space, which is why we may call it \emph{$\rnni$ space}.

\todo{figure: rank + $\nni$ move}

The main purpose of this paper is understanding geometric properties of the $\rnni$ space.
Therefore, further notations need to be introduced.
If every internal node of a tree $T$ has at least one leaf as child, $T$ is called \emph{caterpillar tree}.
A \emph{path} $p$ between trees $T_0$ and $T_m$ in $\rnni$ is a sequence $T_0, T_1, \ldots, T_m$ of trees such that there is an edge $\{T_i, T_{i+1}\}$ for all $i=0,\ldots,m-1$.
For such a path $p$ we say the \emph{length} $|p|$ of $p$ is $m$.
A path $p$ between trees $T$ and $R$ is a \emph{shortest path} if for every path $r$ between $T$ and $R$, $|p| \leq |r|$.
The \emph{distance} $d(T,R)$ between trees $T$ and $R$ is the length of a shortest path between them.
If $p$ is a path only consisting of caterpillar trees we call it \emph{caterpillar path}.
For two caterpillar trees $T$ and $R$ a \emph{shortest caterpillar path} is a caterpillar path that is shortest among all caterpillar paths.
The length of such a shortest caterpillar path is called \emph{caterpillar distance} and denoted by $d_c(T,R)$.
A set $\mathbb S$ of trees is called \emph{convex} if for all pairs of trees $T,R \in \mathbb S$, there exists a shortest path from $T$ to $R$ such that every tree on $p$ is element of $\mathbb S$.

%Is it sufficient to mention what RNNI stands for in the Introduction?

%Restrict to RNNIu (See whether we can get same results for RNNI)

%Definitions inside a tree: cherry, partition, edges represent partitions

%If needed: diameter, radius (in the corresponding chapter?), induced triplet (in that chapter?)

%delete definitions in the following chapters that are already covered here!

%replace rank swap with rank move in the following!!

@@

	\textbf{Rank moves:}
	Within rank moves the labels of internal nodes do not change.
	For an example consider Figure~\ref{fig:rank_swap_internal_label}.

	\begin{figure}[H]
		\centering
		\includegraphics[width=0.8\textwidth]{rank_swap_internal_label}
		\caption{The labelling of the internal nodes that swap ranks (red) do not change within the rank move.
		}
		\label{fig:rank_swap_internal_label}
	\end{figure}

@@

	The task is now to prove that there is always a caterpillar path at least as short as the given path $p$.
	Therefore, we use the following observation:
	Within each $\rnni$ move the ranks of at most two internal nodes change by at most one each.
	This is true for both rank moves and $\nni$ moves, according to our definition of the labelling:
	Within a rank move two nodes exchange their ranks, while there are not necessarily changes of ranks when $\nni$ moves are performed.

@@

%This part is mainly for myself, not sure whether it is important to have it in the paper
The counterexample presented in~\cite{Li1996} works in $\nni$ is based on following observation:
Let $T_1$ and $T_2$ be two caterpillar trees.
If each taxon $a$ is replaced by a cherry $(a_1,a_2)$ in both trees $T_1$ and $T_2$, the resulting trees $T_1'$ and $T_2'$ have distance $d_{\nni}(T_1',T_2') = d_{\nni}(T_1,T_2)$ in the $\nni$ graph.
When proving~\ref{thm:split_nni},~\cite{Li1996} give an example of two trees $T$ and $R$ on $2n$ taxa that consist of two caterpillar trees each that share the root as parent.
The split given by the edges incident to the root is $1, \ldots, n | n+1, \ldots, 2n$ in both trees $T$ and $R$.
For the given example there is a shortest path in $\nni$ where at first $T$ is transformed to a tree $T'$ with $n$ cherries where each cherry contains one taxon of $\{1, \ldots, n\}$ and one taxon of $\{n+1, \ldots, 2n\}$.
Afterwards, $T'$ is tranformed to a tree $R'$ that contains the same cherries as $T'$.
Finally, all cherries of $R'$ are resolved in order to end up in tree $R$.
For proving that this path is shorter than any path preserving the split $1, \ldots, n | n+1, \ldots, 2n$, the fact that $d_{\nni}(T',R') = d_{\nni}(T|_{\{1, \ldots, n\}}, R|_{\{1, \ldots, n\}})$ is being used.
However, this does not work in $\rnni$.
Introducing new cherries means introducing new internal nodes with new ranks.
Since there are some rank moves needed for an $\rnni$ path similar to the shortest $\nni$ path presented above, the path in $\rnni$ that breaks the split $1, \ldots, n | n+1, \ldots, 2n$ is no shortest path.
% How do we assign the ranks to the new cherry nodes?
% We could either create a path where $T'$ and $R'$ have distance equal to $d_{\rnni}(T|_{\{1, \ldots, n\}}, R|_{\{n+1, \ldots, 2n\}})$ and have a lot of additional rank moves before and after that part OR
% Build $T'$ and $R'$ that contain all the cherries within the least number of moves possible (3n-3) and having a longer path between $T'$ and $R'$.
% Also anything in between these two options is possible. Are we able to disprove every option??

@@

% With the knowledge that the set of caterpillar trees is convex in $\rnniu$, we can prove the Cluster Theorem restricted to caterpillar trees.

\begin{conjecture}
    Let $T_1$ and $T_2$ be caterpillar trees.
    If $T_1$ and $T_2$ contain the same cluster $C$, then $C$ is present as cluster in every tree on every shortest path between $T_1$ and $T_2$. \todo{Cluster means non-trivial cluster -> def chapter?}
\end{conjecture}

% Following are some ideas that did not lead to major results so far:
% idea for description of max distance caterpillar trees:

@@

According to Algorithm~\ref{alg:max_dist_tree}, the number of trees with maximum distance to a given tree $T$ depends on the number of cherries of $T$ and the ranks of their internal nodes.
For caterpillar trees it is $(n-1)!$.
The number of caterpillar trees with distance $\Delta(\rnniu)$ from a given caterpillar tree is $2^{n-2}$.

@@

%If this is true, we can follow that radius = diameter
